---
title: "comparing clustering algorithms: dbscan & hdbscan"
author: "Hugo Åkerstrand"
date: "2025-01-02"
output: 
  html_document: 
    keep_md: true
---

# Introduction

This document explores the underlying math of `dbscan` functions `dbscan` and `hdbscan`. 
The goal is to compare the two algorithms and see how they perform on flow cytometry data.

```{r setup, include=TRUE, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(dbscan)
library(flowCore)
data(GvHD)
library(tidyverse)
library(patchwork)
library(tidymodels)
```

Start by loading the data and make it into a tibble with the relevant list columns:

```{r}

gvhd_tibble <- tibble(
  exprs = purrr::map(GvHD, ~ exprs(.x)),                                                            # Expression data
  keywords = purrr::map(GvHD, ~ keyword(.x)),                                                       # Meta data
  exprs_tibble = purrr::map(exprs, function(.x) as_tibble(.x) |> rowid_to_column(var = 'event_id')) # For gating
)

head(gvhd_tibble)
```

For EDA, we will only use the first sample in the data set.

```{r}
test <- gvhd_tibble$exprs_tibble[[1]]

test |> 
   ggplot(aes(x = `FSC-H`, y = `SSC-H`)) +
        geom_hex()
```

First we will look at the hdsbscan code

```{r}
dbscan::hdbscan
```
Let's try some different `minPts` values and see how the clustering changes.
```{r}
minPts <- c(10, 20, 50, 100)
results <- map(minPts, ~dbscan::hdbscan(test, minPts = .x))
plot_list <- map(results, function(.x){
  bind_cols(test, cluster = .x$cluster) |> 
    ggplot(aes(x = `FSC-H`, y = `SSC-H`, color = as.factor(cluster))) +
    geom_point(alpha = 0.2 ) +
    ggtitle(label = paste("minPts =", .x$minPts))
})
wrap_plots(plot_list)
```

Continue with `minPts = 50`

```{r}
hdbscan_50 <- dbscan::hdbscan(test, minPts = 50, gen_hdbscan_tree = TRUE)
plot(hdbscan_50)

```

```{r}

# Recipe for pre-processing the data
mclust_rec <- recipes::recipe(~ `FSC-H` + `SSC-H` + `event_id`, data = test) |>
  update_role(event_id, new_role = 'id variable') |> 
  step_filter(`SSC-H` < 975 & `SSC-H` > 25 & `FSC-H` < 975 & `FSC-H` > 25) |> 
  recipes::step_BoxCox(`FSC-H`, `SSC-H`)

# Prep the recipe on the dbscan filtered data
mclust_prep <- recipes::prep(mclust_rec, test)

# And then call it 
mclust_bake <- recipes::bake(mclust_prep, new_data = NULL)

# And visually inspect the result
mclust_bake |> 
  ggplot(aes(x = `FSC-H`, y = `SSC-H`)) +
    geom_hex() +
    theme_minimal() +
    scale_y_continuous(
      expand = expansion(mult = c(0.1, 0.1)),
      )

```

```{r}
minPts <- c(10, 20, 50, 100)
results <- map(minPts, ~dbscan::hdbscan(mclust_bake, minPts = .x))
plot_list <- map(results, function(.x){
  bind_cols(mclust_bake, cluster = .x$cluster) |> 
    ggplot(aes(x = `FSC-H`, y = `SSC-H`, color = as.factor(cluster))) +
    geom_point(alpha = 0.2 ) +
    ggtitle(label = paste("minPts =", .x$minPts))
})
wrap_plots(plot_list)
```