---
title: "comparing clustering algorithms: dbscan & hdbscan"
author: "Hugo Åkerstrand"
date: "2025-01-02"
output: 
  html_document: 
    keep_md: true
---

# Introduction

Gating single cells is absolutely necessary to avoid the false positive, cells 
aggregates that naturally have a higher fluorescence signal relative to that of 
a single cell. Thus, proper data analysis requires comparison of only single cells. 
This is a simple example of how to gate out single cells using a linear regression model.

Data to be used comes from the `flowAI` package and is simultaneously loaded and
cleaned using `flow_auto_qc`.

```{r setup, include=TRUE, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(dbscan)
library(flowCore)
data(GvHD)
library(tidyverse)
library(patchwork)
```

Start by loading the data and make it into a tibble with the relevant list columns:

```{r}

gvhd_tibble <- tibble(
  exprs = purrr::map(GvHD, ~ exprs(.x)),                                                            # Expression data
  keywords = purrr::map(GvHD, ~ keyword(.x)),                                                       # Meta data
  exprs_tibble = purrr::map(exprs, function(.x) as_tibble(.x) |> rowid_to_column(var = 'event_id')) # For gating
)

head(gvhd_tibble)
```

```{r}
test <- gvhd_tibble$exprs_tibble[[1]]

test |> 
   ggplot(aes(x = `FSC-H`, y = `SSC-H`)) +
        geom_hex()
```

```{r}
minPts <- c(10, 20, 50, 100)
results <- map(minPts, ~dbscan::hdbscan(test, minPts = .x))
plot_list <- map(results, function(.x){
  bind_cols(test, cluster = .x$cluster) |> 
    ggplot(aes(x = `FSC-H`, y = `SSC-H`, color = as.factor(cluster))) +
    geom_point(alpha = 0.2 ) +
    ggtitle(label = paste("minPts =", .x$minPts))
})
wrap_plots(plot_list)
```