---
title: "Sinf cells using a lienar regrssion model"
author: "Hugo Åkerstrand"
date: "2024-12-25"
output: 
  html_document: 
    keep_md: true
---

# Introduction

Gating single cells is absolutely necessary to avoid the false positive, cells 
aggregates that naturally have a higher fluorescence signal relative to that of 
a single cell. Thus, proper data analysis requires comparison of only single cells. 
This is a simple example of how to gate out single cells using a linear regression model.

Data to be used comes from the `flowAI` package and is simultaneously loaded and
cleaned using `flow_auto_qc`.

```{r setup, include=TRUE, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(flowAI)
library(flowCore)
library(tidyverse)
library(patchwork)
library(tidymodels)
data("Bcells")
```
```{r}
b_cleaned <- flowAI::flow_auto_qc(Bcells, html_report = FALSE) # has to be set to FALSE else render fails
```

The data is then transformed into a tibble for easier manipulation.

```{r}
b_tibble <- tibble(
  exprs = purrr::map(b_cleaned, ~ exprs(.x)),                                                            # Expression data
  keywords = purrr::map(b_cleaned, ~ keyword(.x)),                                                       # Meta data
  exprs_tibble = purrr::map(exprs, function(.x) as_tibble(.x) |> rowid_to_column(var = 'event_id'))      # For gating
)

head(b_tibble)
```

For gating singlets we will use the parameters `FSC-A` and `FSC-H`. A visual inspection
of the data shows a linear correlation between the two parameters, as expected.
Note that it is common to first gate on cells (e.g. using `FSC-A` and `SSC-A`), 
but this example on focuses on the `FSC-A` and `FSC-H` gating of singlets.

```{r, echo = FALSE, warning=FALSE}

dotplot_list <- b_tibble$exprs_tibble |> purrr::map(function(.x) {
      ggplot(data = .x, aes(x = `FSC-A`, y = `FSC-H`)) +
        geom_hex()
    }
  )

fsch_list <- b_tibble$exprs_tibble |> purrr::map(function(.x) {
      ggplot(data = .x, aes(x = `FSC-A`)) +
        geom_histogram()
    }
  )

ssch_list <- b_tibble$exprs_tibble |> purrr::map(function(.x) {
      ggplot(data = .x, aes(x = `FSC-H`)) +
        geom_histogram()
    }
  )

wrap_plots(
  c(
    dotplot_list, 
    fsch_list, 
    ssch_list
    ), 
  nrow = 3, 
  guides = 'collect'
)
```

The linear regression model is then trained on the data and used to predict the 
`FSC-A` given `FSC-H`. The model is then used to predict the 80% prediction interval
and keep only these cells. The cells outside the interval are considered to be outliers.

```{r}

linear_spec <- linear_reg() %>% # Create a linear regression model
  set_engine("lm") |>           # Use the lm engine 
  set_mode("regression")        # Set mode to regression

sc_rec <- recipes::recipe(data = b_tibble$exprs_tibble[[1]], `FSC-A` ~ `FSC-H`) |> # Create a recipe where FSC-A is predicted by FSC-H
  step_filter(between(`FSC-A`, 1e03, 2e05), between(`FSC-H`, 1e03, 2e05)) |>       # Filter out extreme values
  step_BoxCox()                                                                    # Transform the data

sc_wflow <- workflow() |> # Create a workflow
  add_recipe(sc_rec) |>   # Add the recipe
  add_model(linear_spec)  # Add the model

sc_fit <- purrr::map(b_tibble$exprs_tibble, function(.x) fit(sc_wflow, .x)) # Fit the model to the data

sc_ci <- purrr::map2(sc_fit, b_tibble$exprs_tibble, function(.x, .y) {
  predict(.x, type = 'pred_int', new_data = .y, level = 0.8) # Predict the 80% prediction interval
})

sc_aug <- purrr::map2(sc_fit, b_tibble$exprs_tibble, function(.x, .y) augment(.x, .y)) |> # Augment the data with the predicted values
  purrr::map2(sc_ci,  ~ bind_cols(.x, .y))                                                # Augment the data with the prediction interval

```

The data is then filtered to only include cells within the 80% prediction interval.
```{r}
sc_aug[[1]] |> select(contains('.pred'), 'FSC-A','FSC-H')                     # Before
sc_aug[[1]] |> dplyr::filter(`FSC-A` >= .pred_lower & `FSC-A` <= .pred_upper) # After
```

Plot the data with the outliers colored.
```{r}

plot_list <- purrr::map(sc_aug, function(.x) {
  .x |>
    mutate(outlier = case_when(
    `FSC-A` < .pred_lower ~ 'outlier',
    `FSC-A` > .pred_upper ~ 'outlier',
    TRUE ~ 'single_cell'
  )) |>
  ggplot(aes(x = `FSC-A`, y = `FSC-H`)) +
  geom_point(aes(color = outlier))
})

wrap_plots(plot_list, nrow = 3, guides = 'collect')
```
